\begin{acknowledgements}


As is always the case in a dissertation, the results this thesis is
a culmination of work and support from multiple people to whom I am
deeply grateful.

First, my academic supervisors, Anders Holst, Henrik Bostr\"{o}m, and Seif Haridi.
They have provided me with guidance, help support throughout my thesis and for
that I am grateful.
My industrial supervisor, Daniel Gillblad, has been as good a lab manager and
research colleague as one can hope. His door was always open and he's helped me
grow as a researcher through our joint work and letting me explore the industry through my
internships. Olof G\"{o}rnerup in many ways taught me how research is done when I
was still starting out, with an eye for detail, creative ideas and an adherence to rigorous
methods.

I want to make special mention of Gianmarco De Fransisci Morales, who
guided me through, from a fledgling PhD student to an independent researcher.
Gianmarco taught me how to insist on the highest standards, provided intuitive
explanations for the most complicated topics, and showed me how one can start
from an idea and transform it to a rigorous paper. This dissertation would
not have been possible without his help.

I've been extremely lucky to work in the industry throughout
my studies and work with people much smarter than me, that I've always learned
from. Boxun Zhang at Spotify, Till Rohrmann at Data Artisans, Hossein Vahabi
at Pandora Radio, and Thomas Drake and Hyunsu Cho at Amazon.
I am deeply grateful for the opportunity to work at all these wonderful
places and for the guidance everyone has provided.

Throughout my studies I've had the chance to talk and learn from some
excellent colleagues here. Paris Carbonne is always around
for a coffee and discussions, be that streaming algorithms or the latest
games. Vasia Kalavri who I'll always look up to for how a PhD should be done.
Anis Nasir for excellent times at conferences and around Stockholm.
Erik Ylip\"{a}\"{a} for our many discussions about everything machine
learning, couldn't have asked for a better office-mate.
Bakr Karali for always finding humor about the life of a southerner
in Sweden. Ian Marsh especially for all the pints, advice, and
psychological support, always there when I needed someone to talk to.
Martin Neumann without whom I would have never had this opportunity,
I'm forever indebted and thankful for the good times over a board game.

Finally, I'm grateful for all my friends. Riccardo and Josh
for PhD life chats and Liene for making sure we talk about other things too.
Kehua, Rado, August, and Jacques, even though you guys left Stockholm I'm grateful for the
good times we had while we were all here. Angelina and Dimitris for unequivocal
support of whatever I chose to do, can't tell you how important that is.
Iraklis and Mitsos for always being there, and keeping me grounded to our
roots.

Last but not least, my parents Vasilis and Erifyli, and my siblings Despina and Kostas.
When everything else goes away, we still have each other.

\end{acknowledgements}
